Something about Performance, 2D Component of NEMO. (TBC)

### 2015-02-23 nemolite2d_perf.tex
More text describing opt. steps. Addition of potential column of profile data for optimised version

\begin{table}[!t]
\caption{The performance profile of the original and \psykal versions
  of nemolite2d (for 2000 time-steps of the $128^{2}$ domain and the
  Intel compiler).}
\label{TABLE_profile}
\centering
\begin{tabular}{l|c|c|c|c}
\hline
Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c}{Vanilla \psykal} \\
        & Time (s) & \%  & Time (s) & \% \\
\hline
Continuity  & 0.10  & 2.7  & 0.33  & 6.6  \\
Momentum    & 3.00  & 84.5 & 4.05  & 79.5 \\
BCs         & 0.16  & 4.6  & 0.29  & 5.7  \\
Time-update & 0.29  & 8.1  & 0.41  & 8.1  \\
\hline
\end{tabular}
\end{table}

@@

@@

In fact, in-lining can have a significant effect on the Intel
compiler's ability to vectorise a loop. Taking the loop that calls the
kernel for the u-component of momentum as an example, before in-lining
the compiler reports:
\begin{verbatim}
LOOP BEGIN at time_step_mod.f90(85,7) 
  inlined into nemolite2d.f90(86,11)
 remark #15335: loop was not vectorized: 
  vectorization possible but seems 
  inefficient. 
 --- begin vector loop cost summary ---
 scalar loop cost: 1307 
 vector loop cost: 2391.000 
 estimated potential speedup: 0.540 
 lightweight vector operations: 1133 
 medium-overhead vector operations: 1 
 vectorized math library calls: 1 
 --- end vector loop cost summary ---
LOOP END
\end{verbatim}
After we have manually in-lined the kernel body, the compiler reports:
\begin{verbatim}
LOOP BEGIN at time_step_mod.f90(97,7) 
  inlined into nemolite2d.f90(86,11)
 LOOP WAS VECTORIZED
 --- begin vector loop cost summary ---
 scalar loop cost: 1253 
 vector loop cost: 521.750 
 estimated potential speedup: 2.350 
 lightweight vector operations: 697 
 medium-overhead vector operations: 1 
 vectorized math library calls: 1 
 --- end vector loop cost summary ---
LOOP END
\end{verbatim}
Looking at the `estimated potential speedup' in the compiler outputs
above, it is clear that the way in which the compiler vectorises the
two versions must be very different. This conclusion is borne out by
the fact that if one persuades the compiler to vectorise the first
version (through the use of a directive) then the performance of the
resulting binary is worse than that where the loop is left
un-vectorised.

The profiling data in Table~\ref{TABLE_profile} shows that it is the
Momentum section that accounts for the bulk of the model run-time.  We
therefore chose to in-line the two kernels (one for each of the $x$
and $y$ components of momentum) called within this section. Given our
findings with the shallow code, we have chosen to do full, manual
inlining of the kernel code.

@@

Further comparison of the diagnostic output from the Cray and Intel
compilers revealed that the Cray compiler was vectorising the
Continuity section while the Intel compiler reported that it was
unable to do so due to dependencies. After some experimentation we
found that this was due to the compiler's limited analysis of the way
components of Fortran derived types were being used.

@@

@@

>>>>

\begin{table}[!t]
\caption{The performance profile of the original and \psykal versions
  of nemolite2d (for 2000 time-steps of the $128^{2}$ domain and the
  Intel compiler).}
\label{TABLE_profile}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla \psykal} &
\multicolumn{2}{c}{Optimised \psykal} \\
        & Time (s) & \%  & Time (s) & \%  & Time(s) & \% \\
\hline
Continuity  & 0.10  & 2.7  & 0.33  & 6.6  & 0.11 & \\
Momentum    & 3.00  & 84.5 & 4.05  & 79.5 & 2.09 & \\
BCs         & 0.16  & 4.6  & 0.29  & 5.7  & 0.27 & \\
Time-update & 0.29  & 8.1  & 0.41  & 8.1  & 0.29 & \\
\hline
\end{tabular}
\end{table}

@@

The profiling data in Table~\ref{TABLE_profile} shows that it is the
Momentum section that accounts for the bulk of the model run-time.  We
therefore chose to attempt to optimise this section first. In-keeping
with the \psykal approach, we are only permitted to optimise the middle
(PSy) layer which, for this section comprises calls to two kernels,
one for each of the $x$ and $y$ components of momentum. These kernels
are relatively large; each comprises roughly 85 lines of Fortran
(excluding variable declarations).

@@

In fact, in-lining can have a significant effect on the Intel
compiler's ability to vectorise a loop. Taking the loop that calls the
kernel for the u-component of momentum as an example, before in-lining
the compiler reports:
\begin{verbatim}
LOOP BEGIN at time_step_mod.f90(85,7) 
  inlined into nemolite2d.f90(86,11)
 remark #15335: loop was not vectorized: 
  vectorization possible but seems 
  inefficient. 
 --- begin vector loop cost summary ---
 scalar loop cost: 1307 
 vector loop cost: 2391.000 
 estimated potential speedup: 0.540 
 lightweight vector operations: 1133 
 medium-overhead vector operations: 1 
 vectorized math library calls: 1 
 --- end vector loop cost summary ---
LOOP END
\end{verbatim}
After we have manually in-lined the kernel body, the compiler reports:
\begin{verbatim}
LOOP BEGIN at time_step_mod.f90(97,7) 
  inlined into nemolite2d.f90(86,11)
 LOOP WAS VECTORIZED
 --- begin vector loop cost summary ---
 scalar loop cost: 1253 
 vector loop cost: 521.750 
 estimated potential speedup: 2.350 
 lightweight vector operations: 697 
 medium-overhead vector operations: 1 
 vectorized math library calls: 1 
 --- end vector loop cost summary ---
LOOP END
\end{verbatim}
Looking at the `estimated potential speedup' in the compiler outputs
above, it is clear that the way in which the compiler vectorises the
two versions must be very different. This conclusion is borne out by
the fact that if one persuades the compiler to vectorise the first
version (through the use of a directive) then the performance of the
resulting binary is worse than that where the loop is left
un-vectorised. {\bf In principle we could investigate this by looking
  at the assembler that the Intel compiler generates\ldots}

For the best possible performance, we have chosen to do full, manual
inlining for the two kernels making up the Momentum section.

@@

Having optimised the Momentum section as much as permitted by the
\psykal approach, we turn our attention to the three remaining
sections of the code. The profile data in Table~\ref{TABLE_profile}
shows that these regions are all comparable in terms of cost. What is
striking however, is that the cost of the Continuity section increases
by more than a factor of three in moving to the \psykal version of the
code.

Comparison of the diagnostic output from the Cray and Intel
compilers revealed that the Cray compiler was vectorising the
Continuity section while the Intel compiler reported that it was
unable to do so due to dependencies. After some experimentation we
found that this was due to limitations in the compiler's analysis of
the way components of Fortran derived types were being used. Each
GOcean field object, in addition to the array holding the local
section of the field, contains a pointer to a GOcean grid object. If a
kernel requires grid-related quantities ({\it e.g.} the grid spacing)
then these are obtained by passing it a reference to the appropriate
array within the grid object. Although these grid-related quantities
are read-only within a compute kernel, if they were obtained from the
same object as that containing the array to which that kernel writes
then the Intel compiler identified a dependency preventing
vectorisation.  This limitation was simply removed by ensuring that
all read-only quantities were accessed via field objects that were
themselves read-only for the kernel at hand.  For instance, the call
to the continuity kernel, which confused the Intel compiler, originally
looked like this:
\begin{verbatim}
do jj = 2, N, 1
 do ji = 2, M, 1

   call continuity_code(ji, jj,      &
                        ssha%data,   &
                        sshn_t%data, &
                        ...,         &
                        ssha%grid%area_t)
\end{verbatim}
where {\it ssha} is the only field that is written to by the kernel.
We remove any potential confusion by instead obtaining the
grid-related (read-only) quantities from a field ({\it sshn\_t} in
this case) that is only read by the kernel:
\begin{verbatim}
do jj = 2, N, 1
 do ji = 2, M, 1

   call continuity_code(ji, jj,      &
                        ssha%data,   &
                        sshn_t%data, &
                        ...,         &
                        sshn_t%grid%area_t)
\end{verbatim}

@@

As with the Momentum kernel, we know that obtaining optimal
performance from both the Gnu and Intel compilers requires that a
kernel be manually inlined at its call site. We do this for the
Continuity kernel in this optimisation step.

@@

Having optimised the Continuity section we finally turn our attention
to the Boundary Condition and Time-Update sections.
